Follow-up Samples

### 2017-02-22 submission/manuscript_outline_20161024.tex
read through edits to streamline Amanda suggestions

\textbf{Results:} The bacterial microbiome changed more in response to
lesion removal in carcinoma cases compared to adenoma cases (P-value
\textless{} 0.05). There was no difference for either the adenoma or
carcinoma group in the relative abundance of any OTU between the pre and
post lesion removal (P-value \textgreater{} 0.05). A model built to
classify lesion had an AUC range of 0.811 - 0.866 while a model built to
classify initial versus follow up samples had an AUC range of 0.641 -
0.805. Follow up samples had a decrease in the positive probability of
lesion (P-value \textless{} 0.05) suggesting a movement towards a more
normal bacterial community. The initial sample model had a decrease in
positive probability for the follow up samples tobe an initial sample
(P-value \textless{} 0.05). The lesion model used a total of 53
variables while the initial sample model used a total of 70 variables. A
total of 23 OTUs were common to both models with the majority of these
classifying to commensal bacteria (e.g. \emph{Bacteroides},
\emph{Clostridiales}, \emph{Blautia}, and \emph{Ruminococcaceae}).

@@

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is currently the third most common cause of cancer
deaths {[}1,2{]}. Over the last few years rate of disease mortality has
seen a significant decrease, thanks mainly to improvements in screening
{[}1{]}. However, despite this improvement there are still approximately
50,000 deaths from the disease per year {[}2{]}.

Over the last few years there has been research into how the bacterial
microbiome could be altered directly on tumor tissue itself {[}3{]}. A
few studies have also shown how this bacterial microbiome {[}4{]} or
specific members within it {[}5{]} could be directly involved with the
pathogenesis of CRC. These studies have helped to provide a tantalizing
link between the bacterial microbiome and CRC. Building off of these
findings, there has been promising work on the bacterial microbiome and
its ability to complement existing screening methods such as Fecal
Immunoglobulin Test (FIT) or act alone as a screening tool {[}6,7{]}.
Although these studies suggest that the bacterial microbiome might
change after treatment there remains limited information on the
bacterial microbiome response to lesion (adenoma or carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome
changes between pre (initial) and post (follow up) removal of lesion
(adenoma or carcinoma). We analyzed changes in alpha and beta diversity
as well as specific Operational Taxonomic Units (OTUs) relative
abundance. We utilized Random Forest to build two models. The first
model was built to classify either lesion and normal samples while the
second was built to classify initial and follow up samples. Subsequent
observations on how these models, as well as specific OTUs within them,
performed pre and post lesion removal helped inform us as to whether
initial and follow up samples were changing and whether it was towards a
more normal bacterial microbiome or not. We then investigated the two
models for similar important OTUs to identify the crucial OTUs for not
only classifying lesion or normal samples but also initial and follow
up.

@@

\textbf{\emph{Bacterial Community and FIT:}} Within our 67 person cohort
we first wanted to test whether there were any broad differences between
initial and follow up samples based on lesion being either adenoma (n =
41) or carcinoma (n = 26). We found that the bacterial community in
those with carcinoma were more dissimilar to their initial sample than
those with adenoma (P-value \textless{} 0.001) {[}Figure 1a{]}. We also
found that there were larger changes in fecal blood (measured by FIT)
for those with carcinoma versus adenoma (P-value \textless{} 0.0001)
{[}Figure 1b{]}. The bacterial community structure before and after
surgery was visualized using NMDS for both adenoma {[}Figure1c{]}
(PERMANOVA \textgreater{} 0.05) and carcinoma {[}Figure 1d{]} (PERMANOVA
\textless{} 0.05). Interestingly, when initial and follow up samples
were compared, regardless of whether the lesions were adenoma or
carcinoma, there was no significant overall difference in beta diversity
(PERMANOVA \textgreater{} 0.05). When investigating alpha diversity
metrics there was no difference found between initial and follow up
samples for lesion, adenoma only, or carcinoma only for any metric
tested {[}Table S1{]}. We also observed that there was no difference in
the relative abundance of any OTU between initial and follow up samples
for lesion, adenoma only, or carcinoma only {[}Figure S1{]}.

@@

\textbf{\emph{Treatment Differences:}} After observing these changes in
the bacterial community and positive probability we wanted to assess
whether additional treatments, such as chemotherapy and radiation, could
have an impact on the results that we observed. There was only a
significant difference for change in positive probability for those
treated with chemotherapy for the initial sample model (P-value
\textless{} 0.05). This suggests that those treated for chemotherapy my
have had a larger change from the initial sample then those without such
treatment. All other variables that were tested showed no difference
based on whether chemotherapy or radiation was used {[}Table S3{]}.

@@

Within our study there was a significant difference for the time elapsed
in the collection of the follow up sample between adenoma and carcinoma
(uncorrected P-value \textless{} 0.05), with time passed being less for
adenoma (253 +/- 41.3 days) than carcinoma (351 +/- 102 days). These
results would indicate that the findings described were specific to the
surgical intervention and that some of the differences observed between
carcinoma and adenoma samples could be due to differences in collection
time between samples for the two different groups. Specifically, it
could confound the observation that carcinomas changed more than
adenomas {[}Figure 1a \& 1d{]}. This confounding though would not affect
the observations where these individuals were grouped together.

@@

Many of the common OTUs between the two models used had many OTUs that
taxonomically classified to potential butyrate producers {[}Table S2{]}.
Another batch of OTUs classified to bacteria that can either degrade
polyphenols or are inhibited by them. Both butyrate and polyphenols are
thought to be protective against cancer in part by reducing inflammation
{[}13{]}. These protective compounds are derived from the breakdown of
fiber, fruits, and vegetables by resident gut microbes. One example of
this potential diet-microbiome-inflammation-polyp axis is that
\emph{Bacteroides}, which was highly prevalent in our models, are known
to be increased in those with high non-meat based protein consumption
{[}14{]}. High protein consumption in general has been linked with an
increased CRC risk {[}15{]}. Conversely, \emph{Bacteroides} are
inhibited by polyphenols which are derived from fruits and vegetables
{[}16{]}. Our data fits with the hypothesis that the microbial
metabolites from breakdown products within our own diet could not only
help to shape the existing community but also have an effect on CRC risk
and disease progression.

@@

Interestingly, within the initial sample model the test data performed
better than the training data. This may have occurred because the
training AUC determined from 20 repeated 10 fold cross validation
removed samples at random and did not take into account that they were
matched samples. Another potential reason is that the model itself may
be over-fit since the total number of samples was not that large.
However, the lesion model did not suffer from these discrepancies.
Further independent studies need to be carried out to verify our
findings and models since we are dealing with correlations that may not
be truly representative of the pathogenesis of disease.

@@

\hypertarget{ref-dejea_microbiota_2014}{}
3. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.

\hypertarget{ref-zackular_manipulation_2016}{}
4. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.

\hypertarget{ref-arthur_microbial_2014}{}
5. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
McCafferty J, et al. Microbial genomic analysis reveals the essential
role of inflammation in bacteria-induced colorectal cancer. Nature
Communications. 2014;5:4724.

>>>>

\textbf{Results:} The bacterial microbiome changed more in response to
lesion removal in carcinoma cases compared to adenoma cases (P-value
\textless{} 0.05). There was no difference for either the adenoma or
carcinoma group in the relative abundance of any OTU between pre and
post lesion removal (P-value \textgreater{} 0.05). A model built to
classify lesion had an AUC range of 0.811 - 0.866 while a model built to
classify initial samples had an AUC range of 0.641 - 0.805. Follow up
samples had a decrease in the positive probability of lesion (P-value
\textless{} 0.05) suggesting a movement towards a more normal bacterial
community. The initial sample model had a decrease in positive
probability for the follow up samples to be an initial sample (P-value
\textless{} 0.05). The lesion model used a total of 53 variables while
the initial sample model used a total of 70 variables. A total of 23
OTUs were common to both models with the majority of these classifying
to commensal bacteria (e.g. \emph{Bacteroides}, \emph{Clostridiales},
\emph{Blautia}, and \emph{Ruminococcaceae}).

@@

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is currently the third most common cause of cancer
deaths {[}1,2{]}. Over the last few years the rate of disease mortality
has seen a significant decrease, thanks mainly to improvements in
screening {[}1{]}. However, despite this improvement there are still
approximately 50,000 deaths from the disease per year {[}2{]}.

Over the last few years a few studies have shown how this bacterial
microbiome {[}3{]} or specific members within it {[}4{]} could be
directly involved with the pathogenesis of CRC. There has also been
research into how the bacterial microbiome could be altered directly on
tumor tissue itself {[}5{]}. These studies have helped to provide a
tantalizing link between the bacterial microbiome and CRC. Building off
of these findings, there has been promising work on the bacterial
microbiome and its ability to complement existing screening methods such
as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool
{[}6,7{]}. Although these studies suggest that the bacterial microbiome
might change after treatment there remains limited information on the
bacterial microbiome response to lesion (adenoma or carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome
changes between pre (initial) and post (follow up) removal of lesion
(adenoma or carcinoma). We analyzed changes in alpha and beta diversity
as well as specific Operational Taxonomic Units (OTUs) relative
abundance. We utilized Random Forest to build two models. The first
model was built to classify lesion and normal samples while the second
was built to classify initial and follow up samples. Subsequent
observations on how these models, as well as specific OTUs within them,
performed pre and post lesion removal helped inform us as to whether
initial and follow up samples were changing and whether it was towards a
more normal bacterial microbiome or not. We also investigated the two
models for similar important OTUs to identify the crucial OTUs for not
only classifying lesion or normal samples but also initial and follow
up.

@@

\textbf{\emph{Bacterial Community and FIT:}} Within our 67 person cohort
we first wanted to test whether there were any broad differences between
initial and follow up samples based on lesion being either adenoma (n =
41) or carcinoma (n = 26). We found that the bacterial community in
those with carcinoma were more dissimilar to their initial sample than
those with adenoma (P-value \textless{} 0.001) {[}Figure 1A{]}. We also
found that there were larger changes in fecal blood (measured by FIT)
for those with carcinoma versus adenoma (P-value \textless{} 0.0001)
{[}Figure 1B{]}. The bacterial community structure before and after
surgery was visualized using NMDS for both adenoma {[}Figure 1C{]}
(PERMANOVA \textgreater{} 0.05) and carcinoma {[}Figure 1D{]} (PERMANOVA
\textless{} 0.05). Interestingly, when initial and follow up samples
were compared, regardless of whether the lesions were adenoma or
carcinoma, there was no significant overall difference in beta diversity
(PERMANOVA \textgreater{} 0.05). When investigating alpha diversity
metrics there was no difference found between initial and follow up
samples for lesion, adenoma only, or carcinoma only for any metric
tested {[}Table S1{]}. We also observed that there was no difference in
the relative abundance of any OTU between initial and follow up samples
for lesion, adenoma only, or carcinoma only {[}Figure S1{]}.

@@

\textbf{\emph{Treatment Differences:}} After observing these changes in
the bacterial community and positive probability we wanted to assess
whether additional treatments, such as chemotherapy and radiation, could
have an impact on the results that we observed. There was only a
significant difference for change in positive probability for those
treated with chemotherapy for the initial sample model (P-value
\textless{} 0.05). This suggests that follow up samples for those
treated for chemotherapy my have had a larger change from the initial
sample then those without such treatment. All other variables that were
tested showed no difference based on whether chemotherapy or radiation
was used {[}Table S3{]}.

@@

Within our study there was a significant difference for the time elapsed
in the collection of the follow up sample between adenoma and carcinoma
(uncorrected P-value \textless{} 0.05), with time passed being less for
adenoma (253 +/- 41.3 days) than carcinoma (351 +/- 102 days). These
results would indicate that the findings described were specific to the
surgical intervention and that some of the differences observed between
carcinoma and adenoma samples could be due to differences in collection
time between samples for the two different groups. Specifically, it
could confound the observation that carcinomas changed more than
adenomas {[}Figure 1A \& 1D{]}. This confounding though would not affect
the observations where these individuals were grouped together {[}Figure
3 \& 4{]}.

@@

Many of the common OTUs between the two models had OTUs that
taxonomically classified to potential butyrate producers {[}Table S2{]}.
Another batch of OTUs classified to bacteria that can either degrade
polyphenols or are inhibited by them. Both butyrate and polyphenols are
thought to be protective against cancer in part by reducing inflammation
{[}13{]}. These protective compounds are derived from the breakdown of
fiber, fruits, and vegetables by resident gut microbes. One example of
this potential diet-microbiome-inflammation-polyp axis is that
\emph{Bacteroides}, which was highly prevalent in our models, are known
to be increased in those with high non-meat based protein consumption
{[}14{]}. High protein consumption in general has been linked with an
increased CRC risk {[}15{]}. Conversely, \emph{Bacteroides} are
inhibited by polyphenols which are derived from fruits and vegetables
{[}16{]}. Our data fits with the hypothesis that the microbial
metabolites from breakdown products within our own diet could not only
help to shape the existing community but also have an effect on CRC risk
and disease progression.

@@

Interestingly, within the initial sample model the test data performed
better than the training data. This may have occurred because the
training AUC determined from 20 repeated 10 fold cross validation
removed samples at random and did not take into account that they were
matched samples. Another potential reason is that the model itself may
be over-fit since the total number of samples was not that large.
However, the lesion model did not suffer from these discrepancies.
Further independent studies need to be carried out to verify our
findings since we are dealing with correlations that may not be truly
representative of the pathogenesis of disease.

@@

\hypertarget{ref-zackular_manipulation_2016}{}
3. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.

\hypertarget{ref-arthur_microbial_2014}{}
4. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
McCafferty J, et al. Microbial genomic analysis reveals the essential
role of inflammation in bacteria-induced colorectal cancer. Nature
Communications. 2014;5:4724.

\hypertarget{ref-dejea_microbiota_2014}{}
5. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.
